### 2023-02-13 chapters/fairness/fvgm/appendix.tex
fvgm appendix

\clearpage
\appendix

\clearpage

		\section{Broader Impact}
		We have presented an efficient framework for verifying fairness of linear classifiers. This work does not bear any negative societal impact. The sole purpose of our framework is to detect bias of machine learning models that are deployed in high-stake decision making.

		\section{Background}

		\subsection{Fairness Metrics}
		In this section, we state the fairness metrics verified in this paper in further details.
		\subsubsection{Group Fairness.} Group fairness is categorized into three families: independence, separation and sufficiency, of which {\framework} verifies independence and separation metrics. 
		The independence metrics state that the predicition of the classifier should be independent of compound sensitive groups. Formally, independence notion specifies an equal PPV across all sensitive groups for a classifier $\alg$, i.e., $\Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  =  \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}'] , \forall \mathbf{a}, \mathbf{a}' \in A$.
		Since satisfying independence exactly is hard, relaxations of independence fairness metrics, such as \textit{disparate impact} and \textit{statistical parity}~\cite{dwork2012fairness,feldman2015certifying}, are proposed.

		\textit{Disparate impact} (DI)~\cite{feldman2015certifying} measures the ratio of PPVs between the most favored group and least favored group, and prescribe it to be close to $1$. Formally, a classifier satisfies $(1 - \epsilon)$-disparate impact if, for $\epsilon \in [0,1] $,
		\[
		\min_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  \ge (1 - \epsilon) \max_{\mathbf{a}'} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}'].
		\]
		Another popular relaxation of independence metrics  is \textit{statistical parity} (SP) that measures the difference of PPV among sensitive groups, and prescribe this to be near zero. Formally, an algorithm satisfies $\epsilon$-statistical parity if, for $\epsilon \in [0,1] $, 
		\[
		\max_{\mathbf{a}, \mathbf{a}'}|\Pr[\hat{Y} =1 | \mathbf{A} = \mathbf{a}] - \Pr [\hat{Y} = 1| \mathbf{A} = \mathbf{a}']| \le \epsilon.
		\]
		For both disparate impact and statistical parity, lower value of $\epsilon$ indicates higher group fairness of the classifier $\alg$.

		In the \textit{separation (or classification parity)} notion of fairness, the predicted label $\hat{Y}$ of a classifier $\alg$ is independent of the sensitive features $\sensitive$ given the class labels $Y$. In case of binary classifiers, a popular separation metric is \textit{equalized odds} (EO)~\cite{hardt2016equality} that computes the difference of false positive rates (FPR) and the difference of true positive rates (TPR) among all compound sensitive groups. 
		Lower value of equalized odds indicates better fairness.
		A classifier $\alg$ satisfies $\epsilon$-equalized odds if, for all compound sensitive groups $\mathbf{a}, \mathbf{a}' \in A$,
		$ \max_{\mathbf{a}, \mathbf{a}'} |\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 0  ] - \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}', Y = 0]| \le \epsilon, $ and $
		\max_{\mathbf{a}, \mathbf{a}'}|\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 1  ] - \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}', Y = 1]| \le \epsilon.
		$

		\subsubsection{Path-specific Causal Fairness.}
		Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\hat{Y} =1 |\mathbf{A}=  \mathbf{a}] $. We consider mediator features $ \mediator \subseteq \nonsensitive $ sampled from the conditional distribution $ {\mathcal{Z}_{|\mathbf{A} = \mathbf{a}_{\max}}} $. This emulates the fact that mediator variables have the same sensitive features $ \mathbf{a}_{\max} $.  For $ \epsilon \in [0,1] $,  path-specific causal fairness is defined as 
		$
		\max_{\mathbf{a}, \mathbf{a}'} |\Pr[\hat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}', \mediator ]| \le \epsilon
		$.

		Therefore, PCF constrains that $ \hat{Y} $ is not directly dependent of $ \sensitive $ while $ \sensitive $ may indirectly affects $ \hat{Y} $ only through $ \mediator $. PCF is a variation of counterfactual fairness and causal fairness without mediator features~\cite{bastani2019probabilistic}.

		\begin{example}
			Following~\cite{bastani2019probabilistic}, we consider a classifier that decides the hiring of employees based on three features: gender (sensitive), years of experience (non-sensitive), and college-participation (mediator). It is practical to consider that gender $ \in $ \{male, female\} can affect the college-participation of individuals, and all three features are determining factors for the hiring process. Let `male' be the most favored group by the classifier, for instance. Path-specific causal fairness (PCF) ensures that a female candidate should be given a job offer with similar probability as a male candidate (by constraining $ \epsilon \approx 0 $). She,  however,  went to (participated in) college as if she were a male candidate while other non-mediator features such as  `years of experience' are the same.  Therefore, PCF measures the effect of gender on job offer, but ignores the effect of gender on whether candidates went to college.
		\end{example}

%		\newpage
		\section{Proofs of Theoretical Results}
		\iffalse
		\begin{lemmarep}
			$ S(B,\tau) $ does not rely on the order of $\mathbf{B}$.\footnote{\blue{This is the rectified version of Lemma~\ref{lm:property-subset-sum}. We note that this correction does not change any claim in Section~\ref{sec:fvgm} except the proposed heuristic in Section~\ref{sec:dp_formulation}, which we address in final version of the paper.}}
		\end{lemmarep}
		\begin{proof}
		The quantifiers in $\mathbf{B}$ can be either universal, existential, or random. The sum of weights for existential and universal variables in $\mathbf{B}$ are calculated as $ w_{\exists} =  \sum_{i|q_i = \exists} \max\{w_i, 0\}$ and  $ w_{\forall} =  \sum_{i|q_i = \forall} \min\{w_i, 0\}  $, respectively by following Eq.~\eqref{eq:dp_recurse}.   Since the contributions of $ w_\exists $ and $ w_\forall $ in $S(B,\tau)$ are fixed, it suffices to prove that $ S(B,\tau) $ does not rely on the order of \textit{random variables} in $\mathbf{B}$.

		Let  $ \tau' =  \tau - w_\exists - w_\forall$ be the revised threshold for random variables $ B' = \{B_i | q_i = \R\} \subseteq\mathbf{B}$. Then $ S(B,\tau) $ is equivalent to solving $ S(B', \tau') $, and we show that $ S(B', \tau') $ does not rely on the order of $ B' $.

		We consider an indicator function $ f_w(i, b) $ that outputs the weight of $ B_i $ based on its assignment $\mathbf{B}$. 
\begin{align*}
		f_w(i, b) = \begin{cases}
	w_i &\text{ if }\mathbf{B}= 1, \\
	0 &\text{ if }\mathbf{B}= 0.
	\end{cases}
\end{align*}
		Let $ \sigma \in \{0,1\}^{|B'|} $ be a vector of  assignments of variables in $ B' $. Then, $ S(B', \tau') $ is computed as
		\[ S(B', \tau') = \sum_{\sigma \in \{0,1\}^{|B'|}}  \mathds{1}[\sum_{i=1}^{|B'|} f_w(i, \sigma_i) \ge \tau'] \prod_{i=1}^{|B'|}\Pr[B_i = \sigma_i]. \]
		 The aforementioned equation represents a stochastic subset-sum problem, where for each assignment $ \sigma $, we compute the decision version of the subset-sum problem corresponding to $ \sigma $ and multiply with probabilities of random-variables. We observe that $ S(B',\tau') $ does not rely on the order of $ B' $, because both multiplication and addition operations are commutative for real numbers. 
			\end{proof}
		\iffalse
		\begin{lemmarep}
			Let $ Q = \{q_i\} $ be the set of quantifiers of $\mathbf{B}$. If $ |Q| = 1$, then $ Q $ is called homogeneous, and otherwise $ Q $ is called heterogeneous. If $ Q $ is homogeneous, then $ S(B,\tau) $ does not rely on the order of $\mathbf{B}$. Otherwise, $ S(B, \tau) $ is order-specific to $\mathbf{B}$.
		\end{lemmarep}

		\begin{proof}
			When $ |Q|  = 1 $, all variables in $\mathbf{B}$ are either existential, universal or random. When all variables are existential, $ S(B,\tau) $ is solved linearly in $ |B| $ as $ S(B, \tau) = \mathds{1}[\sum_i \max\{w_i, 0\} \ge \tau] $. Similarly, when all variables are universal, $ S(B, \tau) = \mathds{1}[\sum_i \min\{w_i, 0\} \ge \tau] $. In both cases,  the sum of $ |B| $ integers does not depend on the order of integers.

			We consider an indicator function $ f_w(i, b) $ that outputs the weight of $ B_i $ depending on its assignment $\mathbf{B}$. 
			\[
			f_w(i, b) = \begin{cases}
			w_i \text{ if }\mathbf{B}= 1 \\
			0 \text{ if }\mathbf{B}= 0
			\end{cases}
			\]
			Let $ \sigma \in \{0,1\}^{|B|} $ be a vector of  assignments of $\mathbf{B}$. When all variables are random, $ S(B, \tau) $ is computed as
			\[ S(B, \tau) = \sum_{\sigma \in \{0,1\}^{|B|}}  \mathds{1}[\sum_{i=1}^{|B|} f_w(i, \sigma_i) \ge \tau] \prod_{i=1}^{|B|}\Pr[B_i = \sigma_i], \]
			which also does not rely on the order of $\mathbf{B}$. Therefore, if $ Q $ is homogeneous, $ S(B,\tau) $ does not rely on the order of $\mathbf{B}$.

			\red{Lemma is not correct}

		\end{proof}
	\fi

		\begin{lemmarep}
			The decision version of the fairness verification problem for linear classifiers is in $\mathrm{NP^{PP}}$.
		\end{lemmarep}

		\begin{proof}

			\textbf{Case 1: Subset-sum problem is $ \mathrm{NP} $-complete by a reduction to $ 3 $-SAT.}

			We show that subset-sum problem is a $ \mathrm{NP} $-complete problem. In particular, we show that computing $ \sum_{i} w_iB_i = \tau $ for $ B_i \in\mathbf{B}$ is $ \mathrm{NP} $-complete through a reduction to $ 3 $-SAT~\cite{kleinberg2006algorithm}. $ 3 $-SAT problem computes a satisfying assignment of  a CNF formula, say $ \phi $, defining over $ n $ variables $ X_1, \dots, X_n $ and $ m $ clauses $ C_1, \dots, C_m $ where each clause has three literals. We start by defining two integer weights $ a_i $ and $ b_i $ associated with each variable $ X_i $, where including $ a_i $ corresponds to assigning $ X_i $ to true and including $ b_i $ corresponds to assigning $ X_i $ to false. We compute $ a_i $ and $ b_i $ in the following. 
			\begin{align*}
				&a_i = 10^{m+i} + \sum_{j|X_i \in C_j} 10^j\\
				&b_i = 10^{m+i} + \sum_{j|\neg X_i \in C_j} 10^j
			\end{align*}

			We construct a subset-sum problem of $ 2n $ variables with weights $ W =  \cup_{i=1}^n\{a_i, b_i\} $. Concretely, let $\mathbf{B}= \{B_{1,1}, B_{1,2}, \dots, B_{n,1}, B_{n,2}\} $ be the set of variables such that $ w(B_{i,1}) = a_i $ and $ w(B_{i,2}) = b_i $ and our goal is to compute an assignment of $\mathbf{B}$ such that $ \sum_{i} w(B_{i,1})B_{i,1} + w(B_{i,2})B_{i,2} = \tau $.

			Let $ \sigma $ be a satisfying assignment of $ 3 $-SAT formula $ \phi $. Given the assignment $\sigma$, we consider a \textit{solution} of the subset-sum problem by assigning $ B_{i, 1} = 1 $ and $ B_{i, 2} = 0 $ whenever $ \sigma(X_i) = 1 $, and $ B_{i, 1} = 0 $ and $ B_{i, 2} = 1 $ whenever $ \sigma(X_i) = 10 $. Intuitively, 
			we consider a subset of weights from $ W $ by including 
			 $ a_i $ if $ X_i $ is true and otherwise we include $ b_i $.  
			 The sum of \textit{included} weights has following properties. 
			\begin{itemize}
				\item  the sum has $ 1 $ in leftmost $ n $ digits, corresponding to $ 10^{m+i} $ for $ i = 1, \dots, n $ as we include one of $ a_i $ or $ b_i $ for each $ i $. 
				\item the sum has $ 1, 2 $, or $ 3 $ in the next $ m $ digits. The digit $ 10^j $ is the number of true literals in clause $ C_j $ and it is non-zero since $ \sigma $ satisfies each clause in  $ \phi $.
				\item $ 0 $ in the rightmost digit.
			\end{itemize}

			In the subset-sum problem, we set the threshold $ \tau = \sum_{i=1}^n 10^{m+i} + 3 \sum_{j=1}^m 10^j $. We additionally consider integer weights $ c_j =  d_j = 10^j $, which will be added to $ W $ as discussed next. For brevity of presentation, we are not assigning Boolean variables in $\mathbf{B}$ for additional weights $ c_j $ and $ d_j $.
			 We then prove that  $ \sum_{i} w(B_{i,1})B_{i,1} + w(B_{i,2})B_{i,2} = \tau $ if and only if $ \phi $ is satisfiable.

			\begin{itemize}
				\item If $ \phi $ is satisfiable, then we select weights $  a _i $ if $ \sigma(X_i) = 1 $, else $ \mathbf{B}_i $. As discussed above, the sum of selected weights has leftmost $ n $ digits of $ \tau $ correct. To make remaining digits correct, we may need to add both of $ c_j $ and $ d_j $, one of $ c_j $ and $ d_j $, or none of $ c_j $ and $ d_j $ depending  if the digit $ 10^j $ in the sum is $ 1,2, $ or $ 3 $. We emphasis that this addition is deterministic as it relies on the number of true literals in $ C_j $ for the assignment $ \sigma $. Thus, for a satisfying assignment of $ 3 $-SAT, there exists a solution of the subset-sum problem.

				\item Finally, we prove that a solution of the subset-sum problem corresponds to a satisfying assignment of $ \phi $.  First we observe that in any digit $ 10^k $, for $ k \le m $, there are at most five weights with $ 1 $ in digit $ 10^k $: the three literals corresponding to clause $ k $ and $ c_k $ and $ d_k $ (otherwise when $ k > m $, there are at most $ 2 $  weights with $ 1 $ in digit $ 10^k $).  Since there are at most $ 5 $ ones in any position, a subset of these weights will not cause any carries in addition. Hence, we can only get a total of $ \tau $ by including the right number of $ 1 $s in any digit. To do this, one needs to include exactly one of $ a_i $ and $ b_i $, and the corresponding truth assignment $ \sigma $ needs to satisfy $ \phi $ so that in each position $ 1 \le j \le m $, we get at least one $ 1 $ in digit $ 10^j $. We can then add $ c_j $ and $ d_j $ to increase the sum of weights to $ \tau $, as discussed previously. This, for a solution of the subset-sum problem, there is a corresponding satisfying assignment of $ 3 $-SAT
			\end{itemize}

			In the above analysis, the construction of $ a_i $ and $ b_i $ has polynomial complexity with respect to $ n $ and $ m $. Therefore, the subset-sum problem $ \sum_{i} w(B_{i,1})B_{i,1} + w(B_{i,2})B_{i,2} = \tau $ and equivalently, $ \sum_{i} w_iB_i = \tau $ is $ \mathrm{NP} $-complete because $ 3 $-SAT is also $ \mathrm{NP} $-complete. We further claim that computing $ \sum_{i} w_iB_i \ge \tau $ is at least as hard as computing $ \sum_{i} w_iB_i = \tau $, which is required for analyzing the complexity of stochastic subset-sum problem, as discussed next.

			\textbf{Case 2: Stochastic subset-sum problem is $ \mathrm{PSPACE} $-complete by a reduction to $ 3 $-SSAT.}

			We observe that a $ 3 $-SSAT formula that is based on a $ 3 $-SAT formula $ \phi $ and an additional specification of quantifiers for each variable in $ \phi $. Concretely, we define a $ 3 $-SSAT formula $$ \Phi \triangleq (\{q(X_i)\}_{i=1}^{n}, \phi) \text{ , where } q(X_i) \in \{\exists, \forall, \R^{p}\} .$$ To analyze the complexity of a stochastic subset-sum problem, we discuss a procedure in the following that is based on a polynomial reduction to $ 3 $-SSAT.

			We apply the aforementioned construction procedure ($ 3 $-SAT to subset-sum problem) to translate  a $ 3 $-SSAT formula to a stochastic sub-set sum problem. In particular, for each quantified variable in $ \phi $, we now discuss the quantifier of variables in the stochastic subset-sum problem.  If $ X_i $ is $ \exists $ (resp.\ $ \forall $) quantified, associated $ B_{i,1} $ and $ B_{i,2} $ are also $ \exists $ (resp.\ $ \forall $) quantified. Moreover, if $ X_i $ is randomized ($ \R^{p_i} $) quantified with $ p_i = \Pr[X_i = 1] $, then both $ B_{i,1} $ and $ B_{i,2} $ are randomized quantified with probability $ \Pr[B_{i,1}] = p_i $ and $ \Pr[B_{i,2}] = 1 - p_i $. Since the decision version of $ 3 $-SSAT is $ \mathrm{PSPACE} $-complete~\cite{littman2001stochastic}  and the presented reduction procedure has polynomial complexity, the decision version of stochastic subset-sum if also $ \mathrm{PSPACE} $-complete.

			\textbf{Case 3: Fairness verification for linear classifiers is in
			$\mathrm{NP^{PP}}$ by a reduction to E-MAJSAT.}

			E-MAJSAT is a specific quantifier arrangement of SSAT problems where existential (or universal) variables appear before random variables. Formally, a E-MAJSAT formula 
			$$ \Phi =  (\{q(X_1) = \exists,\cdots,q(X_i) =\exists, q(X_{i+1}) =\R^{p_{i+1}}, \cdots, q(X_n) = \R^{p_{n}} \}, \phi).$$ The decision version of E-MAJSAT problem has complexity in $\mathrm{NP^{PP}}$~\cite{littman2001stochastic}. Furthermore, the decision version of SSAT problems with universal quantifiers followed by random-quantifiers are also in $\mathrm{NP^{PP}}$ as this a dual problem of E-MAJSAT~\cite{littman2001stochastic}. 
			In Section~\ref{sec:fvgm}, we present fairness verification of linear classifiers defined over Boolean variables using a formulation of stochastic subset-sum problem. Similar to E-MAJSAT,  existential (or universal) variables appear before random variables in our formulation.  Thus, the decision version of fairness verification of linear classifiers is in $\mathrm{NP^{PP}}$.
			\iffalse 
			\red{We further highlight that verification of linear classifiers on continuous real-valued variables is at least as hard as verifying classifiers on Boolean variables.}\fi
		\end{proof}
		\fi

		\begin{lemmarep}
			Let $ n' $ be the number of existential and universal variables in $ \mathbf{B} $. Let $ w_{\exists} = \sum_{B_i \in \mathbf{B} | q_i = \exists} \max\{w_i, 0\}$  and $ w_{\forall} = \sum_{B_i \in \mathbf{B} | q_i = \forall} \min\{w_i, 0\}$ be the considered sum of weights of existential and universal variables, respectively. We can exactly solve {\stochastic} using dynamic programming with time complexity $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall}) + n') $. The space complexity is  $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall})) $.
		\end{lemmarep}

		\begin{proof}
			\textbf{Case 1: All $n$ variables in $\mathbf{B}$ have randomized quantifiers.}

			At step-$i$ of the dynamic programming (Eq.~\eqref{eq:dp_recurse}), we modify the residual threshold of that step, namely $\tau_i$, either by subtracting $w_i$ or by retaining it.
			Now, we observe that the residual threshold $\tau_i$ for any $i \in \lbrace1,\ldots,n\rbrace$ will be in $[0, \tau+|w_{neg}|]$. This holds because if $\tau_i$ crosses these bounds, the dynamic programming is terminated as shown in Equation~\eqref{eq:dp_terminus}.
			Since all weights of $\{w_i\}_{i=1}^{n}$ are integers, the maximum number of values that the residual threshold can take, is $ (\tau + |w_{neg}|) $.
			Thus, we need to store at most $ n(\tau + |w_{neg}|)$ values in the memory for performing dynamic programming with $n$ variables and $ (\tau + |w_{neg}|) $ number of possible weights.
			Thus, the space complexity  is $ \mathcal{O}(n(\tau + |w_{neg}|)) $.

			In order to construct the dynamic programming table, we have to call the $\mathsf{dp}$ function $ \mathcal{O}(n(\tau + |w_{neg}|)) $ times, in the worst-case.
			Thus, the time complexity of our method is $ \mathcal{O}(n(\tau + |w_{neg}|) $.

			\textbf{Case 2: $n'$ variables have existential or universal quantifiers and $n-n'$ variables have randomized quantifiers in $\mathbf{B}$.}

			According to Eq.~\eqref{eq:dp_recurse}, $ w_\exists $ and $ w_\forall $ are the fixed weights of all existential and universal variables, respectively. Therefore, we need to consider at most $ \tau + |w_{neg}| -  w_\exists - w_\forall $ values of residual weights for random variables. By applying analysis in Case 1, the space and time complexity is derived as $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall})) $.

			We note that there is an additional time complexity of $ \mathcal{O}(n') $ for existential and universal variables in Eq.~\eqref{eq:dp_recurse}. Thus the time complexity becomes $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall}) + n') $. We, however, do not require to store any entry for existential and universal variables in $ \mathsf{dp} $  function and thus, the space complexity remains the same as $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall})) $.
		\end{proof}

		\begin{lemmarep}
			Let $ \mathbf{V} \subseteq \mathbf{B} $ be the set of vertices in the Bayesian network and $ n'' $ be the number of existential and universal variables in $ \mathbf{B} \setminus \mathbf{V} $. Let $ w'_{\exists} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | q_i = \exists} \max\{w_i, 0\}$  and $ w'_{\forall} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | q_i = \forall} \min\{w_i, 0\}$ be the sum of considered weights of existential and universal variables, respectively that only appear in $ \mathbf{B} \setminus \mathbf{V} $. To exactly compute {\stochastic} with correlated variables in dynamic programming approach,  time complexity is $ \mathcal{O}(2^{|\mathbf{V}|} + (n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $ and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall})) $.
		\end{lemmarep}

		\input{tables/sanity_check.tex}

		\begin{proof}
			We first separate analysis of space and time complexity for variables in $ \mathbf{V} $ and in $\mathbf{B}\setminus \mathbf{V} $. For each Boolean variable in $ \mathbf{V} $, we enumerate all assignments, which has time complexity of $ 2^{|\mathbf{V}|} $ and there is no space complexity as discussed in Section~\ref{sec:dp_with_BN}.

			For variables in $\mathbf{B}\setminus \mathbf{V} $, we apply analysis from  Lemma~\ref{lemma:complexity_sss}, where we consider $ (n - n'' - |\mathbf{V}|) $ random variables, $ n'' $ existential/universal variables, and residual weights can take at most $ (\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) $ values. Hence, time complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $, and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall})) $

			Combining two cases, overall time complexity is $ \mathcal{O}(2^{|\mathbf{V}|} + (n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $ and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall})) $. 
		\end{proof}

%	\newpage	
	\section{Extended Experimental Evaluations}
	\label{appendix:experiments}
	Each experiment is performed on Intel Xeon E$ 7-8857 $ v$2 $ CPUs with $ 16 $GB memory, $ 64 $bit Linux distribution based on Debian OS and clock speed $ 3 $ GHz. In the following, we discuss extended experimental results.

	\subsection{Accuracy Comparison Among Different Verifiers}
	We have considered a synthetic problem for comparing accuracy among different verifiers. For Example~\ref{example:intro}, we consider `age $ \ge 40 $' as a Bernoulli random  variable with probability $ 0.5 $. For `income' feature ($ I $), we consider two Gaussian distributions $ \Pr[I | \text{age} \ge 40] \sim \mathcal{N}(0.6, 0.1) $ and $ \Pr[I | \text{age} < 40] \sim \mathcal{N}(0.4, 0.1) $ separated by two age groups. Moreover, for `fitness' feature ($ F $), we consider two Gaussian distributions $ \Pr[F | \text{age} \ge 40] \sim \mathcal{N}(0.7, 0.1) $ and $ \Pr[F | \text{age} < 40] \sim \mathcal{N}(0.3, 0.1) $. On this data, the trained LR and SVM classifier has decision boundary as $ 7.26I + 7.4F - 1.34A \ge 6.62 $ and $ 9.37I + 9.75F - 0.34A \ge 9.4 $, respectively.

	In Figure~\ref{fig:synthetic_bn} we show the Bayesian Network on discretized features, in particular for income and fitness features. In Figure~\ref{fig:synthetic_ppv_max_LR} \ to Figure~\ref{fig:synthetic_ppv_min_SVM}, we show PPVs of different classifiers computed by different verifiers, where  {\framework} outputs closest to exactly computed values, in comparison with Justicia. In Figure~\ref{fig:synthetic_sample_size_LR} to Figure~\ref{fig:synthetic_sample_size_SVM}, we show the effect of sample size on {\framework} in measuring fairness metrics: disparate impact and statistical parity, where with increasing sample size, the estimate becomes more accurate.

		\begin{figure}[!t]
		\begin{center}
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Adult_LR}}
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Adult_SVM}}	\\		
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Compas_LR}}
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Compas_SVM}}\\

			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_German_LR}}
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_German_SVM}}		\\
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Titanic_LR}}
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Titanic_SVM}}\\

			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Ricci_LR}}
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Ricci_SVM}}

			\caption{Effect of number of features on the runtime of different datasets for LR and SVM classifiers.}
			\label{fig:time_vary_features}

@@

	In Figure~\ref{fig:time_vary_features}, we present the runtime of different fairness verifiers while varying the number of features in different datasets. We observe that with an increase of features, the runtime increases in general.

	\input{tables/fairness_attacks.tex}
	%\input{tables/verifier_scalability.tex}
	%\input{tables/ssat_vs_subset_sum.tex}
	%\input{tables/fairness_algorithms.tex}
	\input{tables/fairness_metrics_on_compound_sensitive_groups.tex}	
	\input{tables/dag_structure_complexity.tex}

@@

	We show extended results on verifying fairness attack in Figure~\ref{fig:attack_extended} for two fairness metrics: disparate impact (DI) and statistical parity (SP). We observe that {\framework} can detect poisoning attack for both metrics.

	In Figure~\ref{fig:multiple_metrics} we show verification results on compound sensitive groups with respect to multiple fairness metrics. In this figure, we observe that with an increase in the number of groups, fairness metrics worsens\textemdash disparate impact decreases and other three metrics increases. Additionally, in Table~\ref{fig:extended_verifying_fairness_algorithms}, we show extended results for German and COMPAS dataset on verifying fairness algorithms: reweighing and optimized pre-processing algorithms.

	\subsection{Performance Analysis of Bayesian Network}
	In Figure~\ref{fig:results_DAG_complexity}, we analyze the performance of encoding Bayesian Networks of differing complexity. We define the complexity of the network as  $ \frac{|V|}{|\nonsensitive \cup \sensitive|} $, which is the ratio between the number of features appearing in the network and total features. In this figure, as the ratio increases, both computation time of {\framework} and learning time of Bayesian Network increase.

	\begin{table}[!t]

>>>>

\chapter{Fairness Verification with Feature Correlation}
	In this chapter, we discuss the extended experimental results as a continuation of Chapter~\ref{chapter:fvgm}.

	\section{Extended Experimental Results}
	\label{fairness_fvgm_appendix:experiments}
	Each experiment is performed on Intel Xeon E$ 7-8857 $ v$2 $ CPUs with $ 16 $GB memory, $ 64 $bit Linux distribution based on Debian OS and clock speed $ 3 $ GHz.

	\input{tables/fairness/fvgm/sanity_check.tex}

	\subsection{Accuracy Comparison Among Different Verifiers}
	We have considered a synthetic problem for comparing accuracy among different verifiers. For Example~\ref{fvgm_example:intro}, we consider `age $ \ge 40 $' as a Bernoulli random  variable with probability $ 0.5 $. For `income' feature ($ I $), we consider two Gaussian distributions $ \Pr[I | \text{age} \ge 40] \sim \mathcal{N}(0.6, 0.1) $ and $ \Pr[I | \text{age} < 40] \sim \mathcal{N}(0.4, 0.1) $ separated by two age groups. Moreover, for `fitness' feature ($ F $), we consider two Gaussian distributions $ \Pr[F | \text{age} \ge 40] \sim \mathcal{N}(0.7, 0.1) $ and $ \Pr[F | \text{age} < 40] \sim \mathcal{N}(0.3, 0.1) $. On this data, the trained LR and SVM classifier has decision boundary as $ 7.26I + 7.4F - 1.34A \ge 6.62 $ and $ 9.37I + 9.75F - 0.34A \ge 9.4 $, respectively.

	In Figure~\ref{fairness_fvgm_fig:synthetic_bn} we show the Bayesian Network on discretized features, in particular for income and fitness features. In Figure~\ref{fairness_fvgm_fig:synthetic_ppv_max_LR} and Figure~\ref{fairness_fvgm_fig:synthetic_ppv_min_LR}, we show PPVs of logistic regression classifier computed by different verifiers, where  {\fvgm} outputs closest to exactly computed values, in comparison with Justicia. In Figure~\ref{fairness_fvgm_fig:synthetic_sample_size_LR_DI} and Figure~\ref{fairness_fvgm_fig:synthetic_sample_size_LR_SPD}, we show the effect of sample size on {\fvgm} in measuring fairness metrics: disparate impact and statistical parity, where with increasing sample size, the estimate becomes more accurate. Similar observations are recorded for SVM classifier in Figure~\ref{fairness_fvgm_fig:synthetic_results_SVM}.

		\begin{figure}
		\begin{center}
			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Adult_LR}}
			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Adult_SVM}}	\\		
			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Compas_LR}}
			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Compas_SVM}}\\

			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_German_LR}}
			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_German_SVM}}		\\
			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Titanic_LR}}
			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Titanic_SVM}}\\

%			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Ricci_LR}}
%			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Ricci_SVM}}
%			
			\caption{Effect of number of features on the runtime of different datasets for LR and SVM classifiers.}
			\label{fairness_fvgm_fig:time_vary_features}

@@

	In Figure~\ref{fairness_fvgm_fig:time_vary_features}, we present the runtime of different fairness verifiers while varying the number of features in different datasets. We observe that with an increase of features, the runtime increases in general.

	\input{tables/fairness/fvgm/fairness_attacks.tex}
	%\input{tables/fairness/fvgm/verifier_scalability.tex}
	%\input{tables/fairness/fvgm/ssat_vs_subset_sum.tex}
	%\input{tables/fairness/fvgm/fairness_algorithms.tex}
	\input{tables/fairness/fvgm/fairness_metrics_on_compound_sensitive_groups.tex}	
	\input{tables/fairness/fvgm/dag_structure_complexity.tex}

@@

	We show extended results on verifying fairness attack in Figure~\ref{fairness_fvgm_fig:attack_extended} for two fairness metrics: disparate impact (DI) and statistical parity (SP). We observe that {\fvgm} can detect poisoning attack for both metrics. 	In Figure~\ref{fairness_fvgm_fig:multiple_metrics} we show verification results on compound sensitive groups with respect to multiple fairness metrics. In this figure, we observe that with an increase in the number of groups, fairness metrics worsens\textemdash disparate impact decreases and other three metrics increases.

	\subsection{Performance Analysis of Bayesian Network}
	In Figure~\ref{fairness_fvgm_fig:results_DAG_complexity}, we analyze the performance of encoding Bayesian Networks of differing complexity. We define the complexity of the network as  $ \frac{|V|}{|\nonsensitive \cup \sensitive|} $, which is the ratio between the number of features appearing in the network and total features. In this figure, as the ratio increases, both computation time of {\fvgm} and learning time of Bayesian Network increase.

	\begin{comment}
	\begin{table}

	\centering
	\small
	\caption{Extended results for verification of different fairness enhancing algorithms using {\fvgm}. Numbers in bold refer to fairness improvement w.r.t. different fairness metrics. RW and OP refer to reweighing and optimized-preprocessing algorithms respectively.  }
	\label{fairness_fvgm_fig:extended_verifying_fairness_algorithms}
	\setlength{\tabcolsep}{.5em}
